---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  eval      = FALSE
)
```

# ctrest

**ctrest** is an R package for estimating the density of ground-dwelling mammals
and birds from **camera-trap video data** using the
**REST** (Random Encounter and Staying Time) model or the
**RAD-REST** (Random Animal Detection REST) model.

Both models are non-invasive, require no individual recognition, and work with
standard video recordings from fixed camera traps.

---

## Background

The REST model estimates animal density from two key measurements recorded for
every video clip:

1. **Number of passes** — how many times the animal entered the focal area
   (the area clearly visible in the camera's field of view) in that clip.
2. **Staying time** — how long the animal spent inside the focal area.

Combined with **camera trapping effort** (days) and an estimate of the
**proportion of time the animal is active**, the REST model derives density
without requiring a conventional distance-sampling survey (Nakashima et al. 2018).

The **RAD-REST** model is an extension that relaxes the requirement of counting
passes in every video: it is sufficient to count passes in a random subsample,
which greatly reduces the annotation workload (Nakashima et al. 2026).

---

## Installation

```{r install}
# install.packages("devtools")
devtools::install_github("YoshihiroNakashima/ctrest", build_vignettes = TRUE)
```

For a guided tour after installation:

```{r vignette}
library(ctrest)
vignette("ctrest")
```

### System requirements

- R ≥ 4.1
- [NIMBLE](https://r-nimble.org/) is used internally for Bayesian MCMC.
  On Windows, [Rtools](https://cran.r-project.org/bin/windows/Rtools/) must be
  installed. On macOS, Xcode Command Line Tools are required.
- A multi-core CPU is strongly recommended because chains are run in parallel.

---

## Choosing between REST and RAD-REST

| Situation | Recommended model |
|---|---|
| Passes are counted for every video | REST |
| Passes are counted for a random subsample only | RAD-REST |
| You want to minimise annotation time | RAD-REST |
| You want the simplest possible workflow | REST |

REST data can always be analysed with RAD-REST (but not vice versa).

---

## Workflow overview

```
detection_data ──┬── format_station_data() ──── add_effort() ──────────────────┐
                 ├── format_stay()                                               │
                 │         └── bayes_stay_selection()   (optional)              ├──▶ bayes_rest()
                 └── format_activity()                                           │
station_data   ──┘                                                               │
                                                                                 └──▶ bayes_rest_multi()
```

The analysis has five stages:

1. **Format station data** (`format_station_data` + `add_effort`)
2. **Format staying time data** (`format_stay`)
3. **Format activity data** (`format_activity`)
4. *(Optional)* **Select the staying time distribution** (`bayes_stay_selection`)
5. **Estimate density** (`bayes_rest` or `bayes_rest_multi`)

---

## Example datasets

Two example datasets ship with the package:

```{r load-data}
library(ctrest)

head(detection_data)
head(station_data)
```

**`detection_data`** has one row per video clip and contains:

| Column | Description |
|--------|-------------|
| `Station` | Camera station ID |
| `DateTime` | Date and time of detection |
| `Term` | Survey round / visit ID |
| `Species` | Species detected |
| `Nfocal` | Number of passes through the focal area in this clip |
| `Stay` | Staying time (seconds) inside the focal area |
| `Cens` | Censoring flag: `1` = right-censored (animal still in frame at clip end), `0` = observed |

**`station_data`** has one row per camera station and contains:

| Column | Description |
|--------|-------------|
| `Station` | Camera station ID (matches `detection_data$Station`) |
| `x1` | A continuous environmental covariate |
| `x2` | A categorical environmental covariate |

---

## Stage 1: Format station data

`format_station_data()` aggregates the per-video pass counts into
**station-level summaries** and joins station covariates.

For the REST model it produces a column `Y` (total passes per station).
For RAD-REST it additionally produces columns `y_0`, `y_1`, `y_2`, ...
(counts of videos with 0, 1, 2, ... passes observed).

```{r format-station}
# REST
station_data_rest <- format_station_data(
  detection_data   = detection_data,
  station_data     = station_data,
  col_name_station = "Station",
  col_name_species = "Species",
  col_name_y       = "Nfocal",
  model            = "REST"
)

head(station_data_rest)
```

```{r format-station-rad}
# RAD-REST
station_data_rad <- format_station_data(
  detection_data   = detection_data,
  station_data     = station_data,
  col_name_station = "Station",
  col_name_species = "Species",
  col_name_y       = "Nfocal",
  model            = "RAD-REST"
)

head(station_data_rad)
```

### Compute camera trapping effort

`add_effort()` calculates **how many days each camera was operating** by
computing the span between the first and last detection at each station (within
each survey term, if `col_name_term` is supplied).

Set `plot = TRUE` to display a Gantt-style timeline of camera operation — a
useful sanity check before running models.

```{r add-effort}
station_effort_rest <- add_effort(
  detection_data         = detection_data,
  station_data_formatted = station_data_rest,
  col_name_station       = "Station",
  col_name_datetime      = "DateTime",
  col_name_term          = "Term",   # NULL if there are no survey rounds
  plot                   = TRUE
)

head(station_effort_rest)
```

> **Note.** The effort estimate (last detection − first detection) is an
> approximation. For more accurate estimates, provide actual camera operation
> logs directly.

---

## Stage 2: Format staying time data

`format_stay()` joins the detection records to station metadata, renames the
key columns to the standard names expected by the model functions, and flags
any data quality issues.

```{r format-stay}
stay_data <- format_stay(
  detection_data   = detection_data,
  station_data     = station_data,
  col_name_station = "Station",
  col_name_species = "Species",
  col_name_stay    = "Stay",
  col_name_cens    = "Cens"
)

head(stay_data)
```

The output contains `Station`, `Species`, `Stay` (seconds), `Cens` (0/1), and
all covariates from `station_data`. Rows with missing or non-positive staying
times are removed automatically, with a message reporting how many were dropped.

---

## Stage 3: Format activity data

`format_activity()` converts detection datetimes to **radians on a [0, 2π] scale**
so that 00:00 = 0 and 23:59 ≈ 2π. It also filters for independent detections
(one detection per station per species per `indep_time`-minute window).

```{r format-activity}
activity_data <- format_activity(
  detection_data    = detection_data,
  col_name_station  = "Station",
  col_name_species  = "Species",
  col_name_datetime = "DateTime",
  indep_time        = 30   # minimum minutes between independent detections
)

head(activity_data)
```

---

## Stage 4 (optional): Select the staying time distribution

Before running `bayes_rest()`, you can use `bayes_stay_selection()` to compare
candidate probability distributions for staying time via WAIC and choose the
best-fitting one.

Supported distributions: `"exponential"`, `"gamma"`, `"weibull"`, `"lognormal"`.

```{r stay-selection}
stay_sel <- bayes_stay_selection(
  formula_stay       = Stay ~ 1 + x1,   # covariates on mean staying time
  random_effect_stay = NULL,             # or "Station" for a random intercept
  stay_data          = stay_data,
  col_name_cens      = "Cens",
  stay_family        = "lognormal",      # distribution to evaluate
  iter               = 5000,
  warmup             = 1000,
  chains             = 3,
  thin               = 4,
  all_comb           = TRUE,             # compare all covariate subsets
  target_species     = "SP01"
)

print(stay_sel)
```

The printed output shows:

- A WAIC table comparing all candidate models (covariate subsets × random
  effect options), with the best model marked.
- Posterior estimates of the mean staying time from the best model.
- Convergence diagnostics (Rhat).
- A Bayesian p-value assessing model fit.

You can access the raw MCMC samples for further diagnostics:

```{r stay-sel-trace}
MCMCvis::MCMCtrace(stay_sel$samples, pdf = FALSE)
```

---

## Stage 5: Estimate density

### REST model

Pass the data prepared above to `bayes_rest()`. Set `focal_area` to the size of
the camera's focal area in **square metres** — this is the area visible to the
camera within which passes and staying time are measured.

```{r bayes-rest}
result_rest <- bayes_rest(
  formula_stay       = Stay ~ 1 + x1,
  formula_density    = ~ 1,             # density model (no LHS)
  formula_enter      = ~ 1,             # ignored for REST, required for RAD-REST
  station_effort_data = station_effort_rest,
  stay_data          = stay_data,
  activity_data      = activity_data,
  random_effect_stay = NULL,
  activity_estimation = "kernel",       # "kernel" or "mixture"
  bw_adj             = 1.0,            # bandwidth adjustment for kernel estimation
  stay_family        = "lognormal",     # from bayes_stay_selection output
  focal_area         = 1.96,           # focal area in m²
  iter               = 5000,
  warmup             = 1000,
  chains             = 3,
  thin               = 2,
  model              = "REST",
  all_comb           = FALSE,          # TRUE = compare all density covariate subsets
  target_species     = "SP01"
)
```

### RAD-REST model

Use `station_effort_rad` (the RAD-REST formatted station data) and set
`model = "RAD-REST"`. Now `formula_enter` specifies covariates on the mean
number of passes per video.

```{r bayes-rad-rest}
station_effort_rad <- add_effort(
  detection_data         = detection_data,
  station_data_formatted = station_data_rad,
  col_name_station       = "Station",
  col_name_datetime      = "DateTime",
  col_name_term          = "Term"
)

result_rad <- bayes_rest(
  formula_stay        = Stay ~ 1,
  formula_density     = ~ 1,
  formula_enter       = ~ 1,
  station_effort_data = station_effort_rad,
  stay_data           = stay_data,
  activity_data       = activity_data,
  activity_estimation = "kernel",
  stay_family         = "lognormal",
  focal_area          = 1.96,
  iter                = 5000,
  warmup              = 1000,
  chains              = 3,
  thin                = 2,
  model               = "RAD-REST",
  all_comb            = FALSE,
  target_species      = "SP01"
)
```

### Activity estimation: kernel vs mixture

The `activity_estimation` argument controls how the proportion of time the
animal is active is estimated:

| Value | Method | Notes |
|---|---|---|
| `"kernel"` | Fixed kernel density (Rowcliffe et al. 2014) | Fast; `bw_adj` controls bandwidth |
| `"mixture"` | Nonparametric von Mises mixture (Nakashima et al. 2025) | Fully Bayesian; propagates uncertainty; `C` sets max components |

The mixture method is recommended when activity patterns are complex (e.g.,
multimodal) or when you want activity uncertainty to propagate into the density
estimate.

---

## Interpreting the output

```{r print-result}
print(result_rest)
```

The printed summary shows:

```
=== ctrest: Density estimation ===
Species    : SP01
Model      : REST
Stay family: lognormal

--- Model comparison (WAIC) ---
  Model  random_effect_stay   WAIC  Note
  ~ 1    NULL               1234.5  <- best

--- Posterior estimates (best model) ---
  Species  Station  Variable    mean    sd   lower  median   upper  Rhat  n.eff     cv
  SP01     All      density    0.052  0.008  0.038   0.051   0.069  1.00   3000  0.154
  SP01     All      mean_stay  4.230  0.312  3.640   4.218   4.872  1.00   3000  0.074

--- Convergence ---
  All 12 monitored parameter(s): Rhat <= 1.1.

Note: Full MCMC samples : $samples
      Long-format samples: $tidy_samples
      ...
```

- **`density`** (individuals per m²) — multiply by 10 000 to convert to
  individuals per ha, or by 1 000 000 for individuals per km².
- **`mean_stay`** — posterior mean of mean staying time in seconds.
- **`mean_pass`** (RAD-REST only) — posterior mean of mean passes per video.
- **Rhat** — the Gelman-Rubin convergence diagnostic. Values ≤ 1.1 indicate
  good convergence. If Rhat is large, increase `iter` or `chains`.
- **n.eff** — effective sample size. Values ≥ 1000 are generally acceptable.
- **cv** — coefficient of variation (sd / mean).

### Accessing raw MCMC samples

```{r access-samples}
# Full posterior samples (coda::mcmc.list)
result_rest$samples

# Long-format tibble — useful for ggplot2
result_rest$tidy_samples

# Trace plots
MCMCvis::MCMCtrace(result_rest$samples, pdf = FALSE)

# Density plot of a parameter
MCMCvis::MCMCplot(result_rest$samples, params = "density")

# Full posterior summary table
print(result_rest$summary_result)
```

### Activity curve (mixture method only)

When `activity_estimation = "mixture"`, the output includes `$activity_curve`,
a data frame of the posterior activity density function evaluated on a fine time
grid. This can be plotted directly:

```{r plot-activity-curve}
library(ggplot2)

ggplot(result_rest$activity_curve, aes(x = x)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "steelblue", alpha = 0.3) +
  geom_line(aes(y = mean), colour = "steelblue") +
  scale_x_continuous(
    name   = "Time of day",
    breaks = c(0, pi / 2, pi, 3 * pi / 2, 2 * pi),
    labels = c("00:00", "06:00", "12:00", "18:00", "24:00")
  ) +
  ylab("Activity density") +
  theme_bw()
```

### Covariates in the density model

When covariates are included in `formula_density` (or `formula_stay`), one row
per station is returned in `$summary_result`, showing station-specific
estimates:

```{r covariate-example}
result_cov <- bayes_rest(
  formula_stay        = Stay ~ 1,
  formula_density     = ~ 1 + x1,   # density varies with covariate x1
  station_effort_data = station_effort_rest,
  stay_data           = stay_data,
  activity_data       = activity_data,
  activity_estimation = "kernel",
  stay_family         = "lognormal",
  focal_area          = 1.96,
  iter                = 5000,
  warmup              = 1000,
  chains              = 3,
  thin                = 2,
  model               = "REST",
  target_species      = "SP01"
)

# One row per station
print(result_cov$summary_result)
```

Use `all_comb = TRUE` to automatically compare all possible covariate subsets
via WAIC and select the best model:

```{r all-comb}
result_allcomb <- bayes_rest(
  formula_stay        = Stay ~ 1,
  formula_density     = ~ 1 + x1 + x2,   # candidate covariates
  station_effort_data = station_effort_rest,
  stay_data           = stay_data,
  activity_data       = activity_data,
  activity_estimation = "kernel",
  stay_family         = "lognormal",
  focal_area          = 1.96,
  iter                = 5000,
  warmup              = 1000,
  chains              = 3,
  thin                = 2,
  model               = "REST",
  all_comb            = TRUE,             # fit all 4 subsets: ~1, ~x1, ~x2, ~x1+x2
  target_species      = "SP01"
)

# WAIC table for all candidate models
print(result_allcomb$WAIC)
```

---

## Standalone activity estimation

You can also estimate activity patterns independently of density using
`bayes_activity()`. This is useful for comparing activity patterns across
species or seasons, or for quick checks before running the full model.

```{r bayes-activity}
act_result <- bayes_activity(
  activity_data  = activity_data,
  C              = 10,       # maximum number of von Mises components
  iter           = 5000,
  warmup         = 1000,
  chains         = 3,
  thin           = 2,
  target_species = "SP01"
)
```

Print a text summary (no plot):

```{r print-activity}
print(act_result)
```

Print a summary with the activity curve plot:

```{r print-activity-plot}
printResultActivity(act_result, plot = TRUE, bw_adj = 1.0)
```

The plot shows:
- **Blue ribbon**: 95% credible interval of the posterior activity density.
- **Blue line**: Posterior mean activity density.
- **Dashed grey line**: Non-parametric kernel density estimate for comparison.
- **Rug**: Raw detection times.

The activity proportion (probability that the animal is active at a random
moment) is reported in `$summary_result`.

---

## Multi-species density estimation

`bayes_rest_multi()` estimates density for several species simultaneously using
a **hierarchical model** in which species-level random effects are shared
across species, stabilising estimates for rare species.

```{r bayes-rest-multi}
result_multi <- bayes_rest_multi(
  formula_stay        = Stay ~ 1,
  formula_density     = ~ 1,
  formula_enter       = ~ 1,
  station_effort_data = station_effort_rad,   # RAD-REST format
  stay_data           = stay_data,
  activity_data       = activity_data,
  activity_estimation = "kernel",
  stay_family         = "lognormal",
  focal_area          = 1.96,
  iter                = 5000,
  warmup              = 1000,
  chains              = 3,
  thin                = 4,
  model               = "RAD-REST",
  target_species      = c("SP01", "SP02", "SP03")
)

print(result_multi)
```

The output structure is identical to `bayes_rest()`. Density estimates for all
species are returned in `$summary_result`.

---

## Video subsampling for RAD-REST

If you collected REST-format data (passes counted for all videos) but want to
apply RAD-REST to a subsample, use `select_videos()` to draw a balanced random
subsample:

```{r select-videos}
sampled <- select_videos(
  detection_data    = detection_data,
  col_name_species  = "Species",
  col_name_station  = "Station",
  col_name_datetime = "DateTime",
  N_sampled         = 100,        # total videos to select
  Indep_criteria    = 30,         # independence interval in minutes
  target_species    = "SP01",
  seed              = 42
)
```

The function samples videos evenly across stations to avoid over-representing
highly active locations.

---

## Function reference

| Function | Purpose |
|---|---|
| `format_station_data()` | Aggregate pass counts per station; join covariates |
| `add_effort()` | Compute camera trapping effort (days) per station |
| `format_stay()` | Prepare staying time data with censoring |
| `format_activity()` | Convert datetimes to radians; filter independent detections |
| `select_videos()` | Balanced random subsample for RAD-REST |
| `bayes_stay_selection()` | Compare staying time distributions via WAIC |
| `bayes_rest()` | Bayesian density estimation for a single species |
| `bayes_rest_multi()` | Hierarchical Bayesian density estimation for multiple species |
| `bayes_activity()` | Standalone Bayesian activity pattern estimation |
| `printResultActivity()` | Print summary and activity curve plot |

---

## Tips

**My Rhat values are > 1.1.**
Increase `iter` (e.g., to 10 000) and/or `chains` (e.g., to 4). Also check
whether the model is identifiable — highly correlated parameters or very sparse
data can hinder convergence.

**MCMC compilation takes a long time.**
This is normal for NIMBLE; the model is compiled once per chain. Subsequent
runs with the same model structure are not faster, but you can reduce `warmup`
for exploratory runs.

**The activity proportion estimate seems unreasonably high or low.**
Check `$activity_curve`: if the estimated density curve has unrealistic peaks
(e.g., very narrow spikes), try reducing `C` (fewer mixture components) or
using `activity_estimation = "kernel"` instead.

**Density estimates differ between REST and RAD-REST.**
This is expected when the subsample is small. RAD-REST inference is consistent
but has wider uncertainty intervals. With large subsamples the two should agree.

**I want to convert density to abundance.**
Multiply `density` (individuals / m²) by the area of your study region in m².

---

## References

Nakashima, Y., Fukasawa, K., & Samejima, H. (2018). Estimating animal density
without individual recognition using information anchored by camera traps.
*Journal of Applied Ecology*, **55**(2), 900-910.
<https://doi.org/10.1111/1365-2664.13059>

Nakashima, Y. (2021). Potentials and pitfalls of using camera traps for
estimating animal density: a study using REST model across multiple sites and
species. *bioRxiv*.
<https://doi.org/10.1101/2021.05.18.444583>

Nakashima, Y., et al. (2026). RAD-REST: random animal detection model for
density estimation with camera-trap videos.
*Methods in Ecology and Evolution*.
<https://doi.org/10.1111/2041-210x.70248>

Rowcliffe, J. M., Kays, R., Kranstauber, B., Carbone, C., & Jansen, P. A.
(2014). Quantifying levels of animal activity using camera trap data.
*Methods in Ecology and Evolution*, **5**(11), 1170-1179.
<https://doi.org/10.1111/2041-210X.12278>
